Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-06-26 papers/paper-eccentricity/paper-eccentricity.tex
Update references in bib file

\author[0000-0001-5721-9335]{Matias Mannerkoski}
\affiliation{Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland}

@@

%-------------------------- MAIN TEXT STARTS HERE --------------------------%
% INTRODUCTION
\section{Introduction}
Supermassive black holes (SMBHs) are believed to reside at the centres of all massive galaxies \citep[e.g.][]{kormendy2013}.
In the $\Lambda$CDM model as galaxies grow through gas accretion and mergers \citep[e.g.][]{Volonteri2003,naab2017} their SMBHs 
are also expected to interact in a three-phase merger process \citep{begelman1980}.

Firstly, dynamical friction \citep{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scales down to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{Milosavljevic2001,Merritt2013book}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,Rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence \citep[e.g.][]{Mannerkoski2019,Mannerkoski2022}, where the strength of the GW emission strongly depends on the
binary eccentricity \citep{peters1963,peters1964}.

The complex nature of SMBH coalescence in a galaxy merger setting necessitates the use of numerical techniques \citep[e.g.][]{Berentzen2009,Khan2011,Dosopoulou2017,Mannerkoski2019}
to provide quantitative predictions for observational programmes such as ground-based pulsar timing arrays \citep[PTAs, e.g.][]{Lentati2015,Verbiest2016,Arzoumanian2020} and the upcoming Laser Interferometer Space Antenna \citep[LISA, e.g.][]{Amaro-Seoane2023} space mission.
Understanding how faithfully SMBH binary eccentricity is captured in numerical simulations is thus critical in constraining the SMBH merger rate
that the observational GW community aims to detect.

The SMBH binary merging process and its dependence on eccentricity and resolution has been extensively studied using collisionsless merger simulations \citep[e.g.][]{Berentzen2009,Vasiliev2015,Bortolas2016,gualandris2017,gualandris2022}. 
Recently, \citet{nasim2020} have argued that the scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space
discretisation, and that in the real Universe where SMBH masses are far greater than stellar masses $(M_\bullet \gg m_\star)$, SMBH binary eccentricity is a reliably predictable quantity.

@@

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{mannerkoski2023,rantala2017}, in which the dynamics of
SMBH and stars in a small region around them are integrated with an algorithmically regularised integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes Post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{Blanchet2014}.

@@

In numerical simulations with a discretised phase space, the two primary causes of random variation in the impact parameter are the initial particle sampling variation in the merger initial conditions and the Brownian motion of the SMBH binary \citep{merritt2001,Bortolas2016}.
Both of these effects scale as $1/\sqrt{N_{\star,\mathrm{tot}}}$, which can explain the observed scaling of $\sigma_e$ \citep{nasim2020} when the system falls into the region of parameter space where the relation between $e$ and $b$
is approximately linear.
The deviation from this scaling seen in \autoref{fig:convergence} can then be explained by the fact that the eccentricity is not globally a linear function of the impact parameter.

>>>>

\author[0000-0001-5721-9335]{Matias mannerkoski}
\affiliation{Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland}

@@

%-------------------------- MAIN TEXT STARTS HERE --------------------------%
% INTRODUCTION
\section{Introduction}
Supermassive black holes (SMBHs) are believed to reside at the centres of all massive galaxies \citep[e.g.][]{kormendy2013}.
In the $\Lambda$CDM model as galaxies grow through gas accretion and mergers \citep[e.g.][]{volonteri2003,naab2017} their SMBHs 
are also expected to interact in a three-phase merger process \citep{begelman1980}.

Firstly, dynamical friction \citep{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scales down to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{milosavljevic2001,merritt2013}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence \citep[e.g.][]{mannerkoski2019,mannerkoski2022}, where the strength of the GW emission strongly depends on the
binary eccentricity \citep{peters1963,peters1964}.

The complex nature of SMBH coalescence in a galaxy merger setting necessitates the use of numerical techniques \citep[e.g.][]{berentzen2009,khan2011,dosopoulou2017,mannerkoski2019}
to provide quantitative predictions for observational programmes such as ground-based pulsar timing arrays \citep[PTAs, e.g.][]{lentati2015,verbiest2016,arzoumanian2020} and the upcoming Laser Interferometer Space Antenna \citep[LISA, e.g.][]{amaro-seoane2023} space mission.
Understanding how faithfully SMBH binary eccentricity is captured in numerical simulations is thus critical in constraining the SMBH merger rate
that the observational GW community aims to detect.

The SMBH binary merging process and its dependence on eccentricity and resolution has been extensively studied using collisionsless merger simulations \citep[e.g.][]{berentzen2009,vasiliev2015,bortolas2016,gualandris2017,gualandris2022}. 
Recently, \citet{nasim2020} have argued that the scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space
discretisation, and that in the real Universe where SMBH masses are far greater than stellar masses $(M_\bullet \gg m_\star)$, SMBH binary eccentricity is a reliably predictable quantity.

@@

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{mannerkoski2023,rantala2017}, in which the dynamics of
SMBH and stars in a small region around them are integrated with an algorithmically regularised integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes Post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014}.

@@

In numerical simulations with a discretised phase space, the two primary causes of random variation in the impact parameter are the initial particle sampling variation in the merger initial conditions and the Brownian motion of the SMBH binary \citep{merritt2001,bortolas2016}.
Both of these effects scale as $1/\sqrt{N_{\star,\mathrm{tot}}}$, which can explain the observed scaling of $\sigma_e$ \citep{nasim2020} when the system falls into the region of parameter space where the relation between $e$ and $b$
is approximately linear.
The deviation from this scaling seen in \autoref{fig:convergence} can then be explained by the fact that the eccentricity is not globally a linear function of the impact parameter.
